---
title: "APC: Aerobic Plate Count for Serial Dilutions"
author: "Martine Ferguson & John Ihrie"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{APC: Aerobic Plate Count for Serial Dilutions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```


The **MPN** package includes the `apc()` function to estimate the Aerobic Plate
Count (*APC*) point estimate and confidence interval. The *APC* estimates the
average density of colony forming units (CFUs) per milliliter. Although other
techniques exist (*1*) to handle agar plate counts that are
too-numerous-to-count (TNTC), `apc()` uses the maximum likelihood technique of
Haas & Heller (*2*) and Haas et al. (*3*).


## Maximum Likelihood Estimation

As with the Most Probable Number, the Aerobic Plate Count is estimated by
maximizing likelihood. Using a modified version of the notation of Haas et al.
(*3*), we write the likelihood function as:

$$L = \prod_{i=1}^m \frac{{({\lambda}{V_i})} ^ {N_i}}{N_i!}
    {e ^ {-{\lambda}{V_i}}}
    \prod_{j=m+1}^n [1 - \Gamma(N_{L,j}-1, \lambda{V_j})]$$

where

- $\lambda$ is the average microbial density (in CFU/ml) to be estimated
- $m$ is the number of scorable (countable) plates
- $n$ is the total number of plates
- $N_i$ is the CFU count at the $i^{th}$ scorable plate
- $V_i$ is the amount of inoculum (in ml) in the $i^{th}$ scorable plate
- $N_{L,j}$ is the CFU count limit in the $j^{th}$ plate above which the plate
count is considered TNTC
- $V_j$ is the amount of inoculum (in ml) in the $j^{th}$ TNTC plate

As an *R* function:

```{r Define likelihood R function}
L <- function(lambda, count, amount_scor, amount_tntc = NULL, tntc_limit = 100) {
  #likelihood
  scorable <- prod(dpois(count, lambda = lambda * amount_scor))
  if (length(amount_tntc) > 0) {
    incomplete_gamma <- pgamma(tntc_limit - 1, lambda * amount_tntc)
    tntc <- prod(1 - incomplete_gamma)
    return(scorable * tntc)
  } else {
    return(scorable)
  }
}
L_vec <- Vectorize(L, "lambda")
```

`apc()` actually maximizes the log-likelihood function to solve for
$\hat{\lambda}$, the maximum likelihood estimate (MLE) of $\lambda$ (i.e., the
point estimate of *APC*). However, let's demonstrate what is happening in terms
of the likelihood function itself. Assume we start with four plates and 1 ml of
undiluted inoculum. For the first two plates we use a 100-fold dilution; for the
other two plates we use a 1,000-fold dilution. The first two plates were TNTC
with limits of 300 and 250. The other plates had CFU counts of 28 and 20:

```{r Calculate APC estimate}
#APC calculation
library(MPN)
my_count       <- c(28, 20)          #Ni
my_amount_scor <- 1 * c(.001, .001)  #Vi
my_amount_tntc <- 1 * c(.01, .01)    #Vj
my_tntc_limit  <- c(300, 250)        #NLj
(my_apc <- apc(my_count, my_amount_scor, my_amount_tntc, my_tntc_limit))
```

If we plot the likelihood function, we see that $\hat{\lambda}$ maximizes the
likelihood:

```{r Plot likelihood and APC estimate}
my_apc$APC
my_lambda <- seq(29000, 31500, length = 1000)
my_L <- L_vec(my_lambda, my_count, my_amount_scor, my_amount_tntc,
              my_tntc_limit)
plot(my_lambda, my_L, type = "l",
     ylab = "Likelihood", main = "Maximum Likelihood")
abline(v = my_apc$APC, lty = 2, col = "red")
```

If all of the plates have zero counts, the MLE is zero:

```{r All zero counts}
all_zero <- c(0, 0)  #Ni
(apc_all_zero <- apc(all_zero, my_amount_scor)$APC)
my_lambda <- seq(0, 1000, length = 1000)
L_all_zero  <- L_vec(my_lambda, all_zero, my_amount_scor)
plot(my_lambda, L_all_zero, type = "l", ylab = "Likelihood",
     main = "All Zeroes")
abline(v = apc_all_zero, lty = 2, col = "red")
```

## Confidence Intervals

`apc()` computes the confidence interval of $\lambda$ using the likelihood ratio
approach described in Haas et al. (*3*). However, since this approach relies on
large-sample theory, the results are more reliable for larger experiments.

```{r Calculate confidence interval}
my_count       <- c(28, 20)
my_amount_scor <- 1 * c(.001, .001)
my_amount_tntc <- 1 * c(.01, .01)
my_tntc_limit  <- c(300, 250)
(my_apc <- apc(my_count, my_amount_scor, my_amount_tntc, my_tntc_limit))
my_lambda <- seq(25000, 36000, length = 1000)
my_L <- L_vec(my_lambda, my_count, my_amount_scor, my_amount_tntc,
              my_tntc_limit)
plot(my_lambda, my_L, type = "l",
     ylab = "Likelihood", main = "Maximum Likelihood")
abline(v = my_apc$APC, lty = 2, col = "red")
abline(v = my_apc$LB, lty = 3, col = "blue")
abline(v = my_apc$UB, lty = 3, col = "blue")
```

---------

### References

1. Bacteriological Analytical Manual, 8th Edition, Chapter 3,
<https://www.fda.gov/food/foodscienceresearch/laboratorymethods/ucm063346.htm>

2. Haas CN, Heller B (1988). "Averaging of TNTC counts."
*Applied and Environmental Microbiology*, 54(8), 2069-2072.
<https://aem.asm.org/content/54/8/2069>

3. Haas CN, Rose JB, Gerba CP (2014). "Quantitative microbial risk assessment,
Second Ed." *John Wiley & Sons, Inc.*, ISBN 978-1-118-14529-6.
